---
title: "Custom expectations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom expectations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(testthat)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# Pretend we're snapsotting
snapper <- local_snapshotter(fail_on_new = FALSE)
snapper$start_file("snapshotting.Rmd", "test")
```

This vignette shows you how to write your expectations that work identically to the built-in `expect_` functions. 

You can use these either locally by putting them in a helper file, or export them from your package.

## Expectation basics

There are three main parts to writing an expectation, as illustrated by `expect_length()`:

```{r}
expect_length <- function(object, n) {  
  # 1. Capture object and label
  act <- quasi_label(rlang::enquo(object))

  # 2. Verify the expectations
  act_n <- length(act$val)
  if (act_n != n) {
    msg <- sprintf("%s has length %i, not length %i.", act$lab, act_n, n)
    return(fail(msg))
  }
  
  # 3. Pass
  pass(act$val)
}
```

### Capture value and label

The first step in any expectation is to capture the actual object, and generate a label for it to use if a failure occur. All testthat expectations support quasiquotation so that you can unquote variables. This makes it easier to generate good labels when the expectation is called from a function or within a for loop.

By convention, the first argument to every `expect_` function is called `object`, and you capture its value (`val`) and label (`lab`) with `act <- quasi_label(enquo(object))`, where `act` is short for actual (in constrast to expected).

### Verify the expectation

Now we can check if our expectation is met and return `fail()` if not. The most challenging job here is typically generating the error message because you want it to be as self-contained as possible. This means it should typically give both the expected and actual value, along with the name of the object passed to the expectation. testthat expectations use `sprintf()`, but if you're familiar with {glue}, you might want to use that instead.

More complicated expectations will have more `if` statements. For example, we might want to make our `expect_length()` function include an assertion that `object` is a vector:

```{r}
expect_vector_length <- function(object, n) {  
  act <- quasi_label(rlang::enquo(object), arg = "object")

  if (!is.atomic(act$val) || !is.list(act$val)) {
    msg <- sprintf("%s is a %s, not a vector", act$lab, typeof(act$val))
    return(fail(msg))
  }

  act_n <- length(act$val)
  if (act$n != n) {
    msg <- sprintf("%s has length %i, not length %i.", act$lab, act_n, n)
    return(fail(msg))
  }
  
  pass(act$val)
}
```

Note that it's really important to `return(fail())` here. You wont see the problem when interactively testing your function because when run outside of `test_that()`, `fail()` throws an error, causing the function to terminate early. When running inside of `test_that()` however, `fail()` does not stop execution because we want to collect all failures in a given test.

### Pass the test

If no assertions fail, call `pass()` with the input value (usually `act$val`). Returning the input value is good practice since expectation functions are called primarily for their side-effects (triggering a failure). This allows expectations to be chained:

```{r}
mtcars |>
  expect_type("list") |>
  expect_s3_class("data.frame") |> 
  expect_length(11)
```

## Testing your expectations

testthat comes with three expectations designed specifically to test expectations: `expect_success()` and `expect_failure()`:

* `expect_success()` checks that your expectation emits exactly one success and zero failures.
* `expect_failure()` checks that your expectation emits exactly one failure and zero successes. 
* `expect_failure_snapshot()` captures the failure message in a snapshot, making it easier to review if it's useful or not.

```{r}
test_that("expect_length works as expected", {
  x <- 1:10
  expect_success(expect_length(x, 10))
  expect_failure(expect_length(x, 11))

  expect_snapshot_failure(expect_length(x, 11))
})
```
