---
title: "Custom expectations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom expectations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(testthat)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# Pretend we're snapsotting
snapper <- local_snapshotter()
snapper$start_file("snapshotting.Rmd", "test")
```

This vignette shows you how to write your expectations. You can use within your package by putting them in a helper file, or share them with others by exporting them from your package.

## Expectation basics

An expectation has three main parts, as illustrated by `expect_length()`:

```{r}
expect_length <- function(object, n) {  
  # 1. Capture object and label
  act <- quasi_label(rlang::enquo(object), arg = "object")

  # 2. Fail when expectations aren't met
  act_n <- length(act$val)
  if (act_n != n) {
    msg <- sprintf("%s has length %i, not length %i.", act$lab, act_n, n)
    return(fail(msg))
  }
  
  # 3. Pass when expectations are
  pass(act$val)
}
```

The first step in any expectation is to use `quasi_label()` to capture both the value (`$val`) of the first argument and a label (`$lab`) to use failure messages. This is a pattern that exists to support fairly esoteric testthat features; you don't need to understand, just copy and paste it 🙂.

Next, you need to fail, for each way that the `object` violates our expectation. In my experience it's easier to check for problems one by one, because that yields the most informative failure messages. Note that it's really important to `return(fail())` here. You wont see the problem when interactively testing your function because when run outside of `test_that()`, `fail()` throws an error, causing the function to terminate early. When running inside of `test_that()` however, `fail()` does not stop execution because we want to collect all failures in a given test.

Finally, if the object is expected, call `pass()` with the input value (usually `act$val`). Returning the input value is good practice since expectation functions are called primarily for their side-effects (triggering a failure). This allows expectations to be chained:

```{r}
mtcars |>
  expect_type("list") |>
  expect_s3_class("data.frame") |> 
  expect_length(11)
```

## Testing your expectations

testthat comes with three expectations designed specifically to test expectations: `expect_success()` and `expect_failure()`:

* `expect_success()` checks that your expectation emits exactly one success and zero failures.
* `expect_failure()` checks that your expectation emits exactly one failure and zero successes. 
* `expect_failure_snapshot()` captures the failure message in a snapshot, making it easier to review if it's useful or not.

It's important to check that expectations return either one failure or one success because the ensures that reporting is correct. If you 

```{r}
test_that("expect_length works as expected", {
  x <- 1:10
  expect_success(expect_length(x, 10))
  expect_failure(expect_length(x, 11))
})

test_that("expect_length gives useful feedback", {
  x <- 1:10
  expect_snapshot_failure(expect_length(x, 11))
})
```

## Examples

### `expect_vector_length()`

For example, you could imagine a slightly more complex version that first checked if the object was a vector:

```{r}
expect_vector_length <- function(object, n) {  
  act <- quasi_label(rlang::enquo(object))

  if (!is.atomic(act$val) && !is.list(act$val)) {
    msg <- sprintf("%s is a %s, not a vector", act$lab, typeof(act$val))
    return(fail(msg))
  }

  act_n <- length(act$val)
  if (act_n != n) {
    msg <- sprintf("%s has length %i, not length %i.", act$lab, act_n, n)
    return(fail(msg))
  }
  
  pass(act$val)
}
```

To make your failure messages as actionable as possible, state both what the object is and what you expected: 

```{r}
#| error: true
expect_vector_length(mean, 10)
expect_vector_length(mtcars, 15)
```

### `expect_s3_class()`

As another example, imagine if you're checking to see if an object inherits from an S3 class. In R, there's no direct way to tell if an object is an S3 object: you can confirm that it's an object, then that it's not an S4 object. So you might organise your test this way:

```{r}
expect_s3_class <- function(object, class) {
  act <- quasi_label(rlang::enquo(object), arg = "object")

  if (!is.object(act$val)) {
    return(fail(sprintf("%s is not an object.", act$lab)))
  }

  if (isS4(act$val)) {
    return(fail(sprintf("%s is an S4 object, not an S3 object.", act$lab)))
  }

  if (!inherits(act$val, class)) {
    msg <- sprintf(
      "%s inherits from %s not %s.",
      act$lab,
      paste0(class(object), collapse = "/"),
      paste0(class, collapse = "/")
    )
    return(fail(msg))
  }

  pass(act$val)
}
```

```{r}
#| error: true
x1 <- 1:10
TestClass <- methods::setClass("Test", contains = "integer")
x2 <- TestClass()
x3 <- factor()

expect_s3_class(x1, "integer")
expect_s3_class(x2, "integer")
expect_s3_class(x3, "integer")
```

## Repeated code

As you write more expectations, you might discover repeated code that you want to extract out in to a helper. For example, testthat has `expect_true()`, `expect_false()`, and `expect_null()` which are special cases of `expect_equal()`

```{r}
expect_true <- function(object) {
  act <- quasi_label(enquo(object))
  expect_constant_(act, TRUE, ignore_attr = TRUE)
}
expect_false <- function(object) {
  act <- quasi_label(enquo(object))
  expect_constant_(act, FALSE, ignore_attr = TRUE)
}
expect_null <- function(object, label = NULL) {
  act <- quasi_label(enquo(object))
  expect_constant_(act, NULL)
}

expect_constant_ <- function(
  act,
  constant,
  ignore_attr = TRUE,
  trace_env = caller_env()
) {
  comp <- waldo::compare(
    act$val,
    constant,
    x_arg = "actual",
    y_arg = "expected",
    ignore_attr = ignore_attr
  )
  if (length(comp) != 0) {
    msg <- sprintf(
      "%s is not %s\n\n%s",
      act$lab,
      format(constant),
      paste0(comp, collapse = "\n\n")
    )
    return(fail(msg, info = info, trace_env = trace_env))
  }

  pass(act$val)
}
```
