---
title: "third-edition"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{third-edition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(testthat)
```

## waldo

One of the biggest day-to-day differences you'll notice with the 3rd edition is the use of the waldo package inside of `expect_equal()` and `expect_identical()`. The goal of waldo is to find and concisely describe the difference between a pair of R objects, with the primary goal of making it easier to figure out what's gone wrong in your unit tests.

This section provides a brief summary of difference you might hit.

```{r}
local_edition(2)

```

-   environments of functions and formulas ignored ([\#24](https://github.com/r-lib/waldo/issues/24))

-   extptr handling (via proxy?)

-   R6 objects

-   timezones ignored by all.equal

-   tolerance inconsistency between all.equal and compare ([\#30](https://github.com/r-lib/waldo/issues/30))
